Electro-Drop Bouncing in Low-Gravity

### 2018-08-04 doc/paper/aiptemplate.tex
added journal paper draft, updated figures

\documentclass[12pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{fontspec} % This line only for XeLaTeX and LuaLaTeX
\usepackage{pgfplots}
\usepackage{pgf}
\usepackage{bookmark}
\usepackage{color}
\usepackage{tikz}
\graphicspath{ {../figures/} }
%\usepackage{svg}
\usepackage{verbatim}
\usepackage{color,soul}
\usepackage{listings}
\usepackage{setspace}
\usepackage[caption=false]{subfig}
\newlength\figureheight
\newlength\figurewidth
\setlength\figureheight{7cm}
\setlength\figurewidth{10cm}
\usepackage[
backend=biber,
style=phys,
sorting=none
]{biblatex}
\addbibresource{thesis.bib}
\usepackage{import}
\let\pgfimageWithoutPath\pgfimage 
\renewcommand{\pgfimage}[2][]{\pgfimageWithoutPath[#1]{../figures/#2}}
\usepackage{array} 
\usepackage{appendix}
\usepackage{makeidx}
\usepackage{sectsty}
%\chapternumberfont{\large} 
%\chaptertitlefont{\large}
%\allsectionsfont{\normalsize}
\makeindex

\title{Drop Electro-Bouncing in Low-Gravity}
\vspace{-25mm}
\author{Erin S. Schmidt, Mark M. Weislogel}
\date{}

%\usepackage{abstract}
%\renewcommand{\abstractnamefont}{\normalfont\bfseries}
%\renewcommand{\abstracttextfont}{\normalfont\small\itshape}

\raggedbottom
\usepackage{setspace}
\usepackage{afterpage}
\newcommand\blankpage{%
    \null
    \thispagestyle{empty}%
    \addtocounter{page}{-1}%
    \newpage}

\begin{document}

\newcommand{\redline}{\raisebox{2pt}{\tikz{\draw[-,red,solid,line width = 1.5pt](0,0) -- (5mm,0);}}}
%\frontmatter
%\maketitle
%\afterpage{\blankpage}
\begin{titlepage}
\doublespacing

\centering %\vspace*{-10ex}
{\textsf{{Electro-Drop Bouncing in Low-Gravity}}}

\vspace{15ex}
by\\
Erin Stivers Schmidt

\vspace{15ex}
\singlespacing
A thesis submitted in partial fulfillment of the\\
requirements for the degree of

\vspace{11ex}
Master of Science\\
in\\
Mechanical Engineering

\vspace{12ex}
Thesis Committee:\\
Mark Weislogel, Chair\\
Ra\'ul Bayo\'an Cal\\
Gerald Recktenwald\\

\vfill
\vspace{2ex}
Portland State University\\%[8pt]
{\number\year}
\end{titlepage}

%\doublespacing
\clearpage

\frontmatter

\section*{Abstract}
\addcontentsline{toc}{chapter}{\numberline{Abstract}}
\noindent
We investigate the dynamics of spontaneous jumps of water drops from electrically charged superhydrophobic dielectric substrates during a sudden step reduction in gravity level. In the brief free-fall environment of a drop tower, with a non-homogeneous external electric field arising due to dielectric surface charges (with surface potentials ~0.4-1.8 kV), body forces acting on the jumped drops are primarily supplied by polarization stress and Coulombic attraction instead of gravity. This electric body force leads to a drop bouncing behavior similar to well-known phenomena in 1-$g_0$, though occurring for much larger drops ($\sim$ 0.5 mL). We show a simple model for the phenomenon, its scaling, and asymptotic estimates for drop time of flight in two regimes: at short-times close to the substrate when drop inertia balances Coulombic force due to net free charge and image charges in the dielectric substrate and at long-times far from the substrate when drop inertia balances free charge Coulombic force and drag. The drop trajectories are controlled primarily by the dimensionless electrostatic Euler number $\mathbb{E}\mbox{u}$, which is a ratio of inertial to electrostatic forces. To experimentally determine values of $\mathbb{E}\mbox{u}$ we conduct a series of drop tower experiments where we observe the effects of drop volume, net free charge, and static surface potential of the superhydrophobic substrate on drop trajectories. We use a direct search optimization to obtain a Maximum Likelihood Estimate for drop net charge, as we do not measure it directly in experiment. For $\phi \mathbb{E}\mbox{u} /8 \pi > 1$ drops escape the electric field, where $\phi$ is a drop to substrate aspect ratio. However, we do not observe any escapes in our dataset. With an eye towards engineering applications we consider the results in light of the so-called low-gravity phase separation problem with a worked example.

\clearpage
% ========================================

%\vspace*{8ex}
\vspace*{\fill}
%\section*{Dedication}
%\addcontentsline{toc}{chapter}{\numberline{Dedication}}

\emph{To Mom, Laura, and the DDT team}\\
\vspace*{\fill}

\clearpage

% ========================================

%\vspace*{8ex}
\vspace*{\fill}
\section*{Acknowledgements}
\addcontentsline{toc}{chapter}{\numberline{Acknowledgements}}

This work was supported by NASA Cooperative Agreement NNX12A047A. I am also greatly indebted to Andrew Greenberg for his frequent and sage technical advice, and to Joe Shields for his assistance with drop tower experiments. \\ 
\vspace*{\fill}
\clearpage

\tableofcontents
%\afterpage{\blankpage}
\listoffigures
%\afterpage{\blankpage}

%\doublespacing
\mainmatter
\chapter{Introduction}
\begin{quote}
``When the influence of gravity on fluid behavior is diminished or removed, other forces, otherwise of small significance, can assume paramount roles.''
- NRC Report to NASA, 2003 \cite{motil_priorities_2012}
\end{quote}

Our terrestrially born intuition about how liquids flow is easily confounded in a low-gravity environment. This should come as little surprise, as we are creatures evolved at the bottom of a gravity well; gravity is natural to us. However when the magnitude of gravitational body forces become small, other forces come into play in fluid dynamical problems, which are otherwise negligible under normal circumstances in 1-$g_0$, where $g_0=9.81$ m/s$^2$ is the standard gravitational acceleration at sea-level. Our 1-$g_0$ cognitive bias then leads to a plethora of problems which are both simultaneously trivially easy to solve in 1-$g_0$, and which despite decades of study continue to elude solutions in a low-gravity context. One example of these problems is the so-called ``phase separation'' problem of separating a gas phase from a multi-phase flow (or the reverse case), without the aid of buoyancy due to gravity. Sans buoyancy, otherwise mundane activities such as venting a gas or settling a liquid in a tank become problematic \cite{petrash_controlling_1964}. Trapped bubbles can vapor lock life-support, power, or propulsion systems \cite{jenson_passive_2014}. Issues of phase separation have so bedeviled human endeavors in space that an entire Apollo Saturn 1B mission (AS-203) was earmarked to study them \cite{hastings_saturn_1965}. This problem has for some time motivated a quest for substitute body forces, and the present work follows in that well-worn tradition.

A solid beginning to rectifying our problem of poor low-gravity intuition, would be to quantify how wrong our intuition is. First, let us clear up a common misnomer about the state of ``zero-gravity''. The theory of general relativity is based on a fundamental equivalence of gravity and acceleration. By measuring acceleration alone, one cannot determine whether one is in gravity-free space or free-falling in a gravitational field (or conversely whether or not one is uniformly accelerating, or standing at rest in a gravitational field). This is sometimes called the equivalence principle of general relativity (e.g. the equivalence of gravitational and inertial mass). In theoretical terms, a `zero-g' experiment performed in a drop tower on the surface of the Earth is equivalent to the same experiment performed in the nearly gravity-free deep space between galaxies, despite the fact that the drop tower itself is enveloped in a 1-$g_0$ gravity field. At the orbital altitude of the International Space Station (ISS) the gravity level is still nearly 90\% of its value at sea-level on Earth, but the astronauts aboard ISS feel weightlessness because they are in a state of free-fall. In a practical sense true zero-g does not exist anywhere in the universe as there are always small, but nonetheless, measurable forces which can accelerate any massive body. For instance, in low-Earth orbit accelerations due to aerodynamic resistance in the tenuous outer atmosphere are on the order of $1 \times 10^{-6}$ $g_0$ (from hence comes the term ``microgravity''). This small drag acceleration provides some ersatz gravity, but there are also other forces present. Foremost of these are capillary forces.

Capillary forces are due to the cumulative effects of a large number of very short range molecular interactions and in a 1-$g_0$ context are usually quite weak, relatively speaking. However, in a low-gravity environment other forces are so weak that capillary forces become extraordinarily strong in a relative sense. In comparing the relative magnitudes of gravity and the capillary force we arrive at our figure of merit for assessing how wrong our fluid mechanical intuition is likely to be in a low-gravity context. This figure of merit is the Bond number dimensionless group
\[ \mathbb{B}\mbox{o} \equiv \frac{g R^2 \Delta \rho}{\gamma}, 
\]
where $g$ is the acceleration, $R$ is the characteristic interfacial length scale, $\Delta \rho$ is the difference in densities across the interface (which simplifies to $\rho$, for large $\Delta \rho$, as in the case of an air-water interface), and where $\gamma$ is the surface tension. In cases where $\mathbb{B}\mbox{o}$ is small a liquid is effectively in low-gravity regardless of the nominal local gravity level. Very small water drops ($R < 1$ mm), for instance, are in low-gravity for all intents and purposes and have nearly spherical shapes. Contrarily, for cases of large $R$ in a space environment, such as with large diameter spacecraft propellant tanks then $ \mathbb{B}\mbox{o} \gg 1$ even though the acceleration remains small. This is actually a quite useful result, for on the basis of dimensional similarity it is possible to accurately simulate low-gravity fluid mechanics (say of large propellant tanks) even in small drop towers simply by scaling dimensionless groups like the Bond number.

As with capillary forces, electrostatic forces have a multitude of applications as substitutes for gravity. Electrohydrodynamic (EHD) body forces have been suggested to combat thermal stratification of propellants under low-g conditions where buoyancy induced natural convection is very small \cite{blackmon_collection_1965}. In this case electro-convection is an analog of natural convection due to a fluid temperature dependent dielectric permittivity gradient (as opposed to a fluid density gradient) in the presence of an electric body force field. The study also suggested applications in cryogenic tank vent screening and in reducing heat transfer to cryogenic propellants (and thus boiloff losses) by dielectrophoretically positioning ullage vapors around the tank walls as a thermal barrier. EHD forces have been studied for enhancing boiling heat transfer by promotion of bubble detachment and prevention of dryout, again as an ersatz buoyancy \cite{snyder_dielectrophoresis_2001, di_marco_influence_2003, marco_use_2012}. EHD heat pipes, which substitute an electrode structure for the capillary wicking structure of a conventional thermocapillary heat pipe can postpone the wicking limit \cite{jones_electrohydrodynamic_1973}. These are restricted to the use of insulating dielectric liquids, which usually have relatively low thermal conductivity, but are highly robust against dryout, support active flow control, and have low viscous losses. Dielectrophoretic settling of cryogenic propellants, in both total and partial communication Propellant Management Device (PMD) configurations, has been studied analytically in both static \cite{hurwitz_electrohydrodynamic_1966} and dynamic cases \cite{koval_dynamics_1967}, and in drop tower experiments \cite{fax_dielectrophoretic_1969}. EHD forces have also been applied to the low-g slosh baffling problem \cite{boretz_orbital_1970, petrash_use_1968, hurwitz_dielectrophoretic_1968}, and studied extensively for the mitigation of vapor pullthrough during tank draining and concomittent minimization of propellant residuals at burnout in a series of experiments at the NASA Lewis (now Glenn) Research Center 2.2 s drop tower facility \cite{berenyi_dielectrophoretic_1970}.

\section{Spontaneous Drop Jump}
Low-gravity phase separation, drop dynamics, and the dynamics and applications of the so-called spontaneous drop or puddle jump, is an active area of study at the Portland State University Dryden Drop Tower laboratory (DDT). When a nonwetting, gravity-dominated sessile drop (e.g. a puddle), which is initially at rest on a surface in the Cassie-Baxter state suddenly undergoes a large step reduction in $\mathbb{B}\mbox{o}$ it will spontaneously jump away from the surface. The Cassie-Baxter state, sometimes also known as the Fakir state, is a metastable heterogenously non-wetting state on a textured surface characterized by the existence of a vapor-layer that separates roughness features of the surface from the bottom of the drop. The spontaneous drop jump was first observed experimentally in the Soviet Union by Kirko \emph{et al.} \cite{kirko_phenomenon_1970} in 1970 for drops of mercury in hydrochloric acid, and later by Wollman \emph{et al.} in 2016 for water drops in air in a set of experiments conducted using drop towers \cite{wollman_more_2016}. The kinetic energy of the jump is supplied by the defect in free surface energy as the new minimum energy surface equilibrium has approximately constant curvature. If the energy losses by viscous and contact line dynamics are neglected, the available kinetic energy is approximated by
\[KE_2 = SE_2 - SE_1 = [(\gamma A)_{gl} + (\gamma A)_{sg}]_2 - [(\gamma A)_{ls} + (\gamma A)_{gl} + (\gamma A)_{sg}]_1, \]
where $KE$ and $SE$ are the respective kinetic and surface energies, $A$ is the surface area, the subscripts $1$ and $2$ denote the intial 1-$g_0$, and low-g states, and the subscripts ${gl}$, ${sg}$, and ${ls}$ denote the gas-liquid, solid-gas and liquid-solid interfaces respectively. The motive force is due to the inertia of internal flows which occur as the interface deflects under the suddenly lessened $\mathbb{B}\mbox{o}$. The characteristic time scale of the rolling up of the contact line scales as $t_j \sim (\rho V_d/\sigma)^{1/2}$ \cite{attari_puddle_2016}, which resembles the contact time, $\tau \approx 2.6(\rho R^3_d/\sigma)^{1/2}$ presented by Richard \emph{et al.} in 2002 for the related problem of drops impacting hydrophobic surfaces in 1-$g_0$ \cite{richard_surface_2002}. For drops with radial symmetry and sufficiently high initial $\mathbb{B}\mbox{o}$, inward radial capillary waves coalesce at the axis, leading to geysering and creation of satellite droplets by the Rayleigh-Plateau breakup of the geyser. In the case of smaller jumping drops the capillary waves are gradually damped by viscous forces. A time-lapsed composite photograph of spontaneous drop jump can be seen in Figure \ref{fig:nobounce}.
\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{drop_jump}
\caption{The trajectory of a 0.2 mL spontaneous drop jump is captured in a composite image over a period $\sim 0.67$ s presented at $\approx 10$ Hz. \label{fig:nobounce}}
\end{figure}

@@

\section{Electro-Drop Bounce}
During the `rolling-up' of drops under ideal conditions, the spontaneous jump phenomenon is governed by a balance of inertia and surface tension forces, and once aloft the drop motion is nominally in a regime of pure drag. However, other forces can again come into play.

We have observed jumped drops to occasionally decelerate and return to the superhydrophobic surface, rebounding multiple times in the fashion of rigid bodies bouncing under 1-$g_0$. The forces at work in such situations are presumably electrostatic in origin. Not without irony given our earlier discussion of underdeveloped low-gravity intuition, we note that though the electro-drop bounce was an accidental effect during our work on spontaneous drop jump, the phenomenon is interesting itself. This has motivated further study, and the `electro-drop bounce' phenomenon is the subject of this thesis. A time-lapsed composite image showing an example of the phenomenon is shown in Figure \ref{fig:bounce}.
\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{bounce}
\caption{The trajectory of a 0.5 mL drop is captured in a composite image over a single bounce period ($\sim 1.25$ s). The surface potential of the superhydrophobic dielectric in this example is $\varphi_s = 1.25 \pm 0.41$ kV. \label{fig:bounce}}
\end{figure}

@@

There are several possible origins of the electric charge that are ultimately responsible for this phenomenon. It is well known that water acquires positive free charge when in contact with certain polymers, especially polytetrafluoroethylene (PTFE), through a process called contact charging \cite{langmuir_surface_1938}. PTFE, on the other hand, tends to readily acquire negative charge by contact with water. The superhydrophobic surfaces used in the spontaneous drop jump experiments have thin (nanometric) PTFE coatings, and we observe that it is extremely easy to produce significant surface potentials $\varphi_s \sim$ 100-500 V by simply flowing streams of distilled water over them. A study of this water on PTFE contact charging phenomenon was conducted by Yatsuzuka \emph{et al.} \cite{yatsuzuka_electrification_1994}, who suggest that this process results from formation of an electrical double layer driven by selective adsorption of ($\mbox{OH}^-$) ions at the polymer surface; other recent work strongly supports this hypothesis \cite{beattie_intrinsic_2006, strazdaite_water_2015}. Yatsuzuka \emph{et al.} also found that the specific charge on the drops in contact with the surface depends on their both surface velocity and conductivity. The contact charging mechanism is the most likely accidental source of charge on the suphydrophobic surfaces used in our work. Given the large roughness, or the ratio of projected to actual surface area, of the superhydrophobic surfaces used in the experiment, and given that the drops are initially in a Cassie-Baxter state, a somewhat electrically resistive air layer is maintained that reduces grounding of the drops despite the large potential difference between them and the surface charges. If this electrostatic potential is greater than the release of surface energy under the sudden change in $\mathbb{B}\mbox{o}$, in the form of the drop kinetic energy, we expect that the drop will return to an equilibrium state on the charged surface which minimizes the potential.

The source of the net free charge on the drops is another issue. The drop charge could also be due to the contact charging mechanism mentioned previously. For instance, in a 1996 paper, NASA flight engineer Don Pettit discusses the problem of low-gravity flow induced charging of liquids, resulting ultimately from contact charging phenomenon \cite{pettit_donald_flow_????}. Another mechanism for the drop charge is field-induced charging. Field-induced charging occurs due to physical breakup of a conductor having a field-induced dipole (e.g., a physical separation of charge). In our work this might occur when a drop is deposited on the charged surface by a grounded syringe. The metal syringe needle tip, and the liquid in the syringe itself are essentially a ground connection which is broken when the syringe is suddenly removed in the presence of an external electric field. Field-induced charging is at work in the famous Kelvin thunderstorm, and is applied in inkjet, and electrospay technologies, where in each case the breakup is by the Rayleigh-Plateau instability. Notably, in Pettit's aforementioned discussion of contact charging of liquids in low-g, he remarks on accidental electrostatic `hula-ing' of silicone oil drops when ejected from a syringe in the vicinity of a highly-charged polymer surface during an experiment conducted aboard STS-5 by mission specialist Joseph Allen \cite{pettit_donald_flow_????}. Depending on the (highly-variable) electrical conductivity of the silicone oil, and the material of the syringe used in the experiment, the charge could arise just as easily by field-induction as by contact charging. Relatedly, in a series of informal and somewhat whimsical experiments Pettit himself electrostatically orbited small water drops around a triboelectrically charged PTFE knitting needle while aboard ISS during expedition 30/31 \cite{stevenson_electrostatic_2015}. Again, in this case, the drop charge is likely field-induced.

\section{Applications}
This study of the electro-drop bounce phenomenon may have results which can be extended to useful applications in a space environment. External surfaces of spacecraft tend to become charged with time due to the space environment itself. Surface charging largely occurs due to low energy electrons (3-50 keV) which do not penetrate the external surface of the spacecraft \cite{czepiela_charging_1997}. The ultimate sources of these charges are trapped charged particles of the van Allen radiation belts, galactic cosmic rays, and the solar wind. This deposited charge accumulates and can lead to significant potential differences between parts of a spacecraft, sometimes leading to breakdown and discharge called Paschen discharge. Deep dielectric charging occurs when higher energy charged particles penetrate the surface of a dielectric material, which can also lead to large potential differences if the dielectric leakage is lower than the external charging rate. In a fluid mechanical context, this accumulated charge could be problematic during any multiphase venting process, as might be expected to occur during autonomous satellite servicing refueling operations, or during boil-off venting of cryogens stored over long periods in a space environment (as in many crewed Mars mission scenarios, or in cryogenic propellant depot operations). Such venting, if occurring in a region of strong field, will also tend to produce a stream of charged drops, which will be decelerated by the field, causing their eventual return, impacting and potentially contaminating the spacecraft surfaces.

As previously noted, robust phase separation is critical to high reliability multiphase systems used in thermal control and life support. Active solid-state phase separation for life support multiphase flows, especially high void-fraction droplet disperse flows is a possible application of this work. Such flows are encountered in dish washing, laundry, waste solids drying, food processing, the Sabatier CO$_2$ reaction, and possibly in vapor-compression cycle condensers. Phase separation for other droplet disperse flows include electrostatic droplet separators for high-efficiency Rankine-cycle turbines, which require a droplet-free vapor phase entering the turbine, but where conventional superheat approaches come with severe mass penalties \cite{unterberg_zero_1962}. More speculatively, results of this work could be applied in high temperature liquid droplet radiators with electrostatic collection \cite{white_liquid_1987}. Removal of satellite droplets produced during pipetting in wet-lab research outside of a glovebox environment aboard ISS has also been recently suggested as a application of the work [private communication, NASA JSC, E. Unger, 2017]. Drops can become spontaneously charged by contact with standard micropipette tips \cite{choi_spontaneous_2013}, and this free charge can possibly be leveraged for the purposes of phase separation in low-gravity.

\section{Overview of this Thesis}
In this thesis we develop and partly validate a simple model to aid in design of future electrostatic disperse drop phase separators. To this end the work encompasses (1) a mathematical model of the electro-drop bounce, its scale analysis, asymptotic estimates for drop apoapses and times-of-flight, and (2) the results of controlled experiments, where the effect of the key independent variables of drop volume $V_d$ and dielectric surface charge density $\sigma$ on drop trajectories is tested.

>>>>

%% ****** Start of file aiptemplate.tex ****** %
%%
%%   This file is part of the files in the distribution of AIP substyles for REVTeX4.
%%   Version 4.1 of 9 October 2009.
%%
%
% This is a template for producing documents for use with 
% the REVTEX 4.1 document class and the AIP substyles.
% 
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.

%\documentclass[aip,graphicx]{revtex4-1}
\documentclass[aip,reprint, floatfix]{revtex4-1}

\draft % marks overfull lines with a black rule on the right
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgf}
\usepackage{color,soul}
\usepackage{color}
\usepackage{bm}% bold math
\usepackage{tikz}
\usepackage{import}
\let\pgfimageWithoutPath\pgfimage 
\renewcommand{\pgfimage}[2][]{\pgfimageWithoutPath[#1]{../figures/#2}}
\begin{document}

\newcommand{\redline}{\raisebox{2pt}{\tikz{\draw[-,red,solid,line width = 1.5pt](0,0) -- (5mm,0);}}}
\graphicspath{ {../figures/} }

% Use the \preprint command to place your local institutional report number 
% on the title page in preprint mode.
% Multiple \preprint commands are allowed.
%\preprint{}

\title{Electro-Drop Bouncing in Low-Gravity} %Title of paper

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s, 
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command. 
% The \affiliation command should follow the other information.

\author{Erin S. Schmidt}
%\email[]{Your e-mail address}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{Portland State University}

\author{Mark M. Weislogel}
\affiliation{Portland State University}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
% insert abstract here

We investigate the dynamics of spontaneous jumps of water drops from electrically charged superhydrophobic dielectric substrates during a sudden step reduction in gravity level. In the brief free-fall environment of a drop tower, with a non-homogeneous external electric field arising due to dielectric surface charges (with surface potentials ~0.4-1.8 kV), body forces acting on the jumped drops are primarily supplied by polarization stress and Coulombic attraction instead of gravity. This electric body force leads to a drop bouncing behavior similar to well-known phenomena in 1-$g_0$, though occurring for much larger drops ($\sim$ 0.5 mL). We show a 1D model for the phenomenon, its scaling, and asymptotic estimates for drop time of flight in two regimes: at short-times close to the substrate when drop inertia balances Coulombic force due to net free charge and image charges in the dielectric substrate and at long-times far from the substrate when drop inertia balances free charge Coulombic force and drag. The drop trajectories are controlled primarily by the dimensionless electrostatic Euler number $\mathbb{E}\mbox{u}$, which is a ratio of inertia to electrostatic force. To experimentally determine values of $\mathbb{E}\mbox{u}$ we conduct a series of drop tower experiments where we observe the effects of drop volume, net free charge, and static surface potential of the superhydrophobic substrate on drop trajectories. We use a direct search optimization to obtain a Maximum Likelihood Estimate for drop net charge, as we do not measure it directly in experiment. For $\phi \mathbb{E}\mbox{u} /8 \pi > 1$ drops escape the electric field, where $\phi$ is a substrate to drop aspect ratio. However, we do not observe any escapes in our dataset.
\end{abstract}

\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

% Body of paper goes here. Use proper sectioning commands. 
% References should be done using the \cite, \ref, and \label commands
\section{Spontaneous Drop Jump}
When a nonwetting, gravity-dominated sessile drop (e.g. a puddle), which is initially at rest on a surface in the Cassie-Baxter state suddenly undergoes a large step reduction in $\mathbb{B}\mbox{o} \equiv \frac{g R^2 \Delta \rho}{\gamma}$ it will spontaneously jump away from the surface, where $g$ is the acceleration, $R$ is the characteristic interfacial length scale, $\Delta \rho$ is the difference in densities across the interface (which simplifies to $\rho$, for large $\Delta \rho$, as in the case of an air-water interface), and where $\gamma$ is the surface tension. The spontaneous drop jump was first observed experimentally in the Soviet Union by Kirko \emph{et al.} \cite{kirko_phenomenon_1970} in 1970 for drops of mercury in hydrochloric acid, and later by Wollman \emph{et al.} in 2016 for water drops in air in a set of experiments conducted using drop towers \cite{wollman_more_2016}. The kinetic energy of the jump is supplied by the defect in free surface energy as the new minimum energy surface equilibrium has approximately constant curvature. The motive force is due to the inertia of internal flows which occur as the interface deflects under the suddenly lessened $\mathbb{B}\mbox{o}$. For drops with radial symmetry and sufficiently high initial $\mathbb{B}\mbox{o}$, inward radial capillary waves coalesce at the axis, leading to geysering and creation of satellite droplets by the Rayleigh-Plateau breakup of the geyser. In the case of smaller jumping drops the capillary waves are gradually damped by viscous forces.

@@

During the `rolling-up' of drops under ideal conditions, the spontaneous jump phenomenon is governed by a balance of inertia and surface tension forces, and once aloft the drop motion is nominally in a regime of pure drag. However, other forces can come into play.

We have observed jumped drops to occasionally decelerate and return to the superhydrophobic surface, rebounding multiple times in the fashion of rigid bodies bouncing under 1-$g_0$. The forces at work in such situations are presumably electrostatic in origin. A time-lapsed composite image showing an example of the phenomenon is shown in Figure \ref{fig:bounce}, and schematically in Figure \ref{fig:bounce_time}.
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{bounce}
\caption{The trajectory of a 0.5 mL drop is captured in a composite image over a single bounce period ($\sim 1.25$ s) presented at $\approx 10$ Hz. The surface potential of the superhydrophobic dielectric is $\varphi_s = 1.25 \pm 0.41$ kV. \label{fig:bounce}}
\end{figure}

\begin{figure}[htb]
\centering
\resizebox{0.5\textwidth}{!}{\input{../figures/bounce_series.pgf}}
\caption{The trajectory time-series of a 0.05 mL drop with surface potential $\varphi_s = 0.75 \pm 0.5$ kV. \label{fig:bounce_time}}
\end{figure}

@@

We expect that if the drop electrostatic potential energy is greater than the release of surface energy under the sudden change in $\mathbb{B}\mbox{o}$, in the form of the drop kinetic energy, the drop will return to an equilibrium state on the charged surface which minimizes the potential.

What is the origin of the electric field arising on the superhydrophobic surfaces? It is well known that water acquires positive free charge when in contact with certain polymers, especially polytetrafluoroethylene (PTFE), through a process called contact charging \cite{langmuir_surface_1938}. PTFE, on the other hand, tends to readily acquire negative charge by contact with water. The superhydrophobic surfaces used in the spontaneous drop jump experiments have thin (nanometric) PTFE coatings, and we observe that it is extremely easy to produce significant surface potentials $\varphi_s \sim$ 100-500 V by simply flowing streams of distilled water over them. A study of this water on PTFE contact charging phenomenon was conducted by Yatsuzuka \emph{et al.} \cite{yatsuzuka_electrification_1994}, who suggest that this process results from formation of an electrical double layer driven by selective adsorption of ($\mbox{OH}^-$) ions at the polymer surface; other recent work supports this hypothesis \cite{beattie_intrinsic_2006, strazdaite_water_2015}. Given the large roughness, or the ratio of projected to actual surface area, of the superhydrophobic surfaces used in the experiment, and given that the drops are initially in a Cassie-Baxter state, a somewhat electrically resistive air layer is maintained that reduces grounding of the drops despite the large potential difference between them and the surface charges.

The source of the net free charge on the drops is another issue. The drop charge could be due to the contact charging mechanism mentioned previously. For instance, in a 1996 paper, NASA flight engineer Don Pettit discusses the problem of low-gravity flow induced charging of liquids, resulting ultimately from contact charging phenomenon \cite{pettit_donald_flow_????}. However, a more likely mechanism for the drop charge is field-induced charging. Field-induced charging occurs due to breakup of a conductor having a field-induced dipole (e.g., a physical separation of charge). In our work this might occur when a drop is deposited on the charged surface by a grounded syringe. The metal syringe needle tip, and the liquid in the syringe itself are essentially a ground connection which is broken when the syringe is suddenly removed in the presence of an external electric field. Field-induced charging is at work in the famous Kelvin thunderstorm, and is applied in inkjet, and electrospay technologies, where in each case the breakup is by the Rayleigh-Plateau instability. Notably, in Pettit's aforementioned discussion of contact charging of liquids in low-g, he remarks on accidental electrostatic `hula-ing' of silicone oil drops when ejected from a syringe in the vicinity of a highly-charged polymer surface during an experiment conducted aboard STS-5 by mission specialist Joseph Allen \cite{pettit_donald_flow_????}. Depending on the (highly-variable) electrical conductivity of the silicone oil, and the material of the syringe used in the experiment, the charge could arise just as easily by field-induction as by contact charging. Relatedly, in a series of informal and somewhat whimsical experiments Pettit himself electrostatically orbited small water drops around a triboelectrically charged PTFE knitting needle while aboard ISS during expedition 30/31 \cite{stevenson_electrostatic_2015}. Again, in this case, the drop charge is likely field-induced.

\section{Theory}
\subsection{Equation of Motion}
We develop here a simple 1-dimensional model of the dynamics of drops dominated by electrostatic forces. We treat a drop as a particle with radius $R_d$, which translates vertically along the central axis of a charged dielectric square sheet substrate with initial velocity $U_0$. The equation of motion for this system is given by,
\begin{equation}
m y'' = - F_D - F_E,
\label{gov_eqn}
\end{equation}
subject to
\begin{equation}
y(0) = R_d, \hspace{5 mm} \mbox{and} \hspace{5 mm} y'(0) = U_0,
\end{equation}
where $m$ is the drop mass, $y'' = \frac{d^2 y}{d t^2}$ is the drop acceleration, $\mathbf{F}_D$ is the drag force which always opposes motion, and $\mathbf{F}_E$ is the electrostatic force. The assumed initial conditions are such that, when $\mathbb{B}\mbox{o}$ is suddenly reduced at the start of the drop tower free-fall period, the drop jumps with instantaneous initial velocity $U_0$ from its 1-$g_0$ resting position with fixed radius $R_d$ at $t=0$. The dynamical system is sketched schematically Figure \ref{apparatus}. We now define models for each of the forces in this equation.

\begin{figure}[ht]
\centering
\includegraphics[width=0.45\textwidth]{../figures/apparatus0.pdf}
\label{apparatus}
\caption{Schematic representation of drop jump with return and rebound from an electrically charged superhydrophobic substrate. The characteristic time and length scales $t_c$ and $y_c$ describe the time of flight and apoapse associated with the drop trajectory.}
\end{figure}

For the intermediate range of Reynolds numbers $1 \leq \mathbb{R}\mbox{e} \equiv \frac{2UR_d}{\nu} \leq 2000 $ observed in our experiments, we assume the force of drag acting on the drop to be quadratic.

In modeling the electrostatic force we begin with the standard electrohydrodynamic (EHD) approximation \cite{saville_electrohydrodynamics:_1997}. Under a DC electric field, we assume that the real part of the dielectric permittivity $\epsilon$, $\mbox{Re} \left( \epsilon \right) \approx  \mbox{constant}$. We also assume that electric currents are small enough that the effects of magnetic fields can be neglected. The validity of this assumption rests on the characteristic time scale $\tau_e = \epsilon /\sigma_e \ll 1$, where $\tau_e$ is the ratio of absolute dielectric permittivity $\epsilon = \kappa \epsilon_0$, to conductivity $\sigma_e$, of the medium, $\kappa$ is the relative dielectric permittivity, and $\epsilon_0$ is the vacuum permittivity. This characteristic time $\tau_e$ is also known as the relaxation time, and is a measure of how quickly the polarization of a dielectric responds to a change in electric field. Given the conductivity and permittivity in the limiting case of extremely-pure water ($ \epsilon \approx 80$, $\sigma_e = 18.2 \times 10^{6}$ $\Omega\mbox{cm}$) \cite{yatsuzuka_electrification_1994}, we estimate $\tau_e \approx 4 \times 10^{-6}$ s. The relaxation time for the common distilled water that is actually used in the drop experiments is undoubtedly shorter due to the presence of solvated ions. Neglecting the effects of an electric double layer on hydration of ions in the water or the ambient atmosphere due to the relatively massive size of the drops studied, the assumption of small relaxation time further implies that the free charge present in the drops will remain approximately constant during the typical time interval of a low-gravity experiment.

Supposing that electrical forces acting on free charges and dipoles in a fluid are transferred directly to the fluid itself, the overall electrical body force will be the divergence of the Maxwell stress tensor $\tau_m $,
\begin{eqnarray} \label{e_force}
 \mathbf{F}_E &=& \nabla \cdot \tau_m \nonumber \\ 
 &=& \rho_f \mathbf{E} + \frac{1}{2} \left| E \right|^2 \nabla \epsilon - \nabla \left( \frac{1}{2} \rho \left( \frac{\partial \epsilon}{\partial \rho} \right)_T \left| E \right|^2 \right) .
\end{eqnarray}
The first term on the right hand side of this expression is the well known Coulombic force or electrophoretic force, which arises from the presence of free charge in an external electric field. We expect this term to dominate the electric force in a DC field. The second term is the force arising from polarization stresses due to a nonuniform field acting across a gradient in permittivity. This force is widely termed the dielectrophoretic (DEP) force. The third term describes forces due to electrostriction. It has been noted by Melcher and Hurwitz \cite{hurwitz_electrohydrodynamic_1966} that the electrostriction term is the gradient of a scalar and can thus be canonically lumped together with the hydrostatic pressure for incompressible fluids. We neglect it in our analysis.
